Davos: a Python package ``smuggler'' for constructing lightweight reproducible notebooks

### 2023-09-10 paper/main.tex
WIP rewriting projects section

For greater control over the behavior of \texttt{smuggle} statements, Davos
defines an additional construct called the ``onion comment.'' An onion comment
is a special type of inline comment that may be placed on a line containing a
\texttt{smuggle} statement to customize how Davos searches for the smuggled
package locally and, if necessary, downloads and installs it. Onion comments
follow a simple format based on the ``type comment'' syntax introduced in PEP
484~\cite{vanREtal14}, and are designed to make managing packages with Davos
intuitive and familiar. To construct an onion comment, users provide the name
of the installer program (e.g., \texttt{pip}) and the same arguments one would
use to manually install the package as desired via the command line:
\begin{center}
  \includegraphics[width=0.9\textwidth]{figs/snippet1}
\end{center}
Occasionally, a package's distribution name (i.e., the name used
when installing it) may differ from its top-level module name (i.e., the name
used when importing it). In such cases, an onion comment may be used to ensure
that Davos installs the proper package if it cannot be found locally:
\begin{center}
  \includegraphics[width=0.9\textwidth]{figs/snippet2}
\end{center}
Because onion comments may be constructed to specify any aspect of
the installer program's behavior, they provide a mechanism for precisely controlling
how, where, and when smuggled packages are installed. Critically, if an onion
comment includes a version specifier~\cite{CoghStuf13}, Davos will ensure that
the version of the package loaded into the notebook matches the specific
version requested, or satisfies the given version constraints. If the smuggled
package exists locally, Davos will extract its version information from its
metadata and compare it to the specifier provided. If the two are incompatible
(or no local installation is found), Davos will download, install, and load a
suitable version of the package instead:
\begin{center}
  \includegraphics[width=0.9\textwidth]{figs/snippet3}
\end{center}
Onion comments can also be used to \texttt{smuggle} specific VCS references (e.g.,
Git~\cite{TorvHama05} branches, commits, tags, etc.):
\begin{center}
  \includegraphics[width=0.9\textwidth]{figs/snippet4}
\end{center}
Davos processes onion comments internally before forwarding arguments to the
installer program. In addition to preventing shared notebooks from executing
arbitrary code in a user's shell, this enables Davos to adapt its behavior
based on how particular flags will affect the behavior of the installer
program. For example, including \texttt{pip}'s \texttt{--no-input} flag will also
temporarily enable Davos's non-interactive mode (see Sec.~\ref{subsec:config}).
Similarly, if an onion comment contains either the \texttt{-I}/\texttt{--ignore-installed},
\texttt{-U}/\texttt{--upgrade}, or \texttt{--force-reinstall} flag, Davos will
skip checking for a local copy of the smuggled package before installing a
new one:
\begin{center}
  \includegraphics[width=0.9\textwidth]{figs/snippet5}
\end{center}
Since the purpose of an onion comment is to describe how a smuggled package
should be installed (if necessary) so that it can be loaded and used
immediately, options that would cause the package not to be installed (such as
\texttt{-h}/\texttt{--help} or \texttt{--dry-run}) are disallowed. Additionally,
when using a Davos project to isolate smuggled packages (the default behavior;
see Sec.~\ref{subsec:projects}), onion comments may not contain options that
would change the package's installation location (such as
\texttt{-t}/\texttt{--target}, \texttt{--root}, or \texttt{--prefix}). However, if
the user disables project-based isolation and specifies \texttt{--target <dir>},
Davos will ensure that \texttt{<dir>} is included in the module search path (i.e.,
\texttt{sys.path}), prepending it if necessary, so the package can be loaded.

@@

To prevent Davos-enhanced notebooks from having unwanted side-effects on the user's environment, Davos automatically isolates packages installed via \texttt{smuggle} statements using a custom scheme called ``projects.'' Functionally, a Davos project is similar to a standard Python virtual environment (e.g., created with the standard library's \texttt{venv} module or a third-party tool like \texttt{virtualenv}~\cite{BickEtal07}): it consists of a directory (within a hidden \texttt{.davos} folder in the user's home directory) that houses third-party packages needed for a particular project or task. However, unlike standard virtual environments, Davos projects do not need to be manually activated and deactivated, do not contain separate Python or \texttt{pip} executables, and \textit{extend} the user's main Python environment rather than replace it.

When Davos is imported into a notebook, a notebook-specific project directory is automatically created (if it does not exist already). 
%When Davos is imported into a notebook, a notebook-specific project directory is automatically created (if it does not exist already), named for the absolute path to the notebook file.

Notebook-specific projects are named for the absolute path to the notebook file.

>>>>

For greater control over the behavior of \texttt{smuggle} statements, Davos
defines an additional construct called the ``onion comment.'' An onion comment
is a special type of inline comment that may be placed on a line containing a
\texttt{smuggle} statement to customize how Davos searches for the smuggled
package locally and, if necessary, downloads and installs it. Onion comments
follow a simple format based on the ``type comment'' syntax introduced in PEP
484~\cite{vanREtal14}, and are designed to make managing packages with Davos
intuitive and familiar. To construct an onion comment, users provide the name
of the installer program (e.g., \texttt{pip}) and the same arguments one would
use to manually install the package as desired via the command line:
\begin{center}
  \includegraphics[width=0.9\textwidth]{figs/snippet1}
\end{center}
Occasionally, a package's distribution name (i.e., the name used
when installing it) may differ from its top-level module name (i.e., the name
used when importing it). In such cases, an onion comment may be used to ensure
that Davos installs the proper package if it cannot be found locally:
\begin{center}
  \includegraphics[width=0.9\textwidth]{figs/snippet2}
\end{center}
Because onion comments may be constructed to specify any aspect of
the installer program's behavior, they provide a mechanism for precisely controlling
how, where, and when smuggled packages are installed. Critically, if an onion
comment includes a version specifier~\cite{CoghStuf13}, Davos will ensure that
the version of the package loaded into the notebook matches the specific
version requested (or satisfies the given version constraints). If the smuggled
package exists locally, Davos will extract its version information from its
metadata and compare it to the specifier provided. If the two are incompatible
(or no local installation is found), Davos will download, install, and load a
suitable version of the package instead:
\begin{center}
  \includegraphics[width=0.9\textwidth]{figs/snippet3}
\end{center}
Onion comments can also be used to \texttt{smuggle} specific VCS references (e.g.,
Git~\cite{TorvHama05} branches, commits, tags, etc.):
\begin{center}
  \includegraphics[width=0.9\textwidth]{figs/snippet4}
\end{center}
Davos processes onion comments internally before forwarding arguments to the
installer program. In addition to preventing shared notebooks from executing
arbitrary code in a user's shell, this enables Davos to adjust its behavior
based on how particular flags will affect the behavior of the installer
program. For example, including \texttt{pip}'s \texttt{--no-input} flag will also
temporarily enable Davos's non-interactive mode (see Sec.~\ref{subsec:config}).
Similarly, if an onion comment contains either \texttt{-I}/\texttt{--ignore-installed},
\texttt{-U}/\texttt{--upgrade}, or \texttt{--force-reinstall}, Davos will
install and load a new copy of the smuggled package without first checking
for it locally:
\begin{center}
  \includegraphics[width=0.9\textwidth]{figs/snippet5}
\end{center}
Since the purpose of an onion comment is to describe how a smuggled package
should be installed (if necessary) so that it can be loaded and used
immediately, options that would normally cause the package not to be installed
(such as \texttt{-h}/\texttt{--help} or \texttt{--dry-run}) are disallowed. Additionally,
when using a Davos ``project'' to isolate smuggled packages (the default behavior;
see Sec.~\ref{subsec:projects}), onion comments may not contain options that
would change the package's installation location (such as
\texttt{-t}/\texttt{--target}, \texttt{--root}, or \texttt{--prefix}). However, if
the user disables project-based isolation and specifies \texttt{--target <dir>},
Davos will ensure that \texttt{<dir>} is included in the module search path (i.e.,
\texttt{sys.path}), prepending it if necessary, so the package can be loaded.

@@

%To prevent Davos-enhanced notebooks from having unwanted side-effects on the user's environment, Davos automatically isolates packages installed via \texttt{smuggle} statements using a custom, virtual environment-like system called ``projects.''
%Functionally, Davos projects are similar to standard Python virtual environments (e.g., created with the standard library's \texttt{venv} module or a third-party tool like \texttt{virtualenv}~\cite{BickEtal07}), but are lighter-weight and simpler to use. 
%Like a standard virtual environment, a Davos project consists of a directory (within a hidden \texttt{.davos} folder in the user's home directory) that houses third-party packages needed for a particular Python-based project \textcolor{red}{[confusing reuse of ``project''?]}, workflow, or task. 
%However, unlike standard virtual environments, Davos projects do not need to be manually created, activated, or deactivated; do not contain separate Python or \texttt{pip} executables; and function to \textit{extend} the user's main Python environment rather than replace it.
To prevent Davos-enhanced notebooks from having unwanted side-effects on the user's environment, any packages installed via \texttt{smuggle} statements are automatically isolated using a custom, virtual environment-like system called ``projects.'' 
Davos projects are similar to standard Python virtual environments (e.g., created with the standard library's \texttt{venv} module or a third-party tool like \texttt{virtualenv}~\cite{BickEtal07}) with a few important differences. Like a standard virtual environment, a Davos project consists of a directory (within a hidden \texttt{.davos} folder in the user's home directory) that houses third-party packages needed to run a particular Python program, workflow, or task. However, unlike standard virtual environments, Davos projects do not need to be manually created, activated, or deactivated, and 
%function to \textit{extend} 
serve to \textit{supplement}
the user's existing Python environment rather than replace it.
%To prevent Davos-enhanced notebooks from having unwanted side-effects on the user's environment, any packages installed via \texttt{smuggle} statements are automatically isolated using a custom, virtual environment-like system called ``projects.'' 
%Davos projects are similar to standard Python virtual environments (e.g., created with the standard library's \texttt{venv} module or a third-party tool like \texttt{virtualenv}~\cite{BickEtal07}): a project consists of a directory (within a hidden \texttt{.davos} folder in the user's home directory) that houses third-party packages needed to run a particular Python program, workflow, or task. However, unlike standard virtual environments, Davos projects do not need to be manually created, activated, or deactivated, and 
%%function to \textit{extend} 
%serve to \textit{supplement}
%the user's existing Python environment rather than replace it.

When Davos is imported into a notebook, a project directory specific to that notebook is automatically created (if it does not exist already). Any packages (or specific package versions) smuggled into that notebook that are not already available in the notebook's main runtime environment will be installed into the notebook's project directory, along with any dependencies of those packages not available in the main environment.

Additionally, when a \texttt{smuggle} statement in that notebook is executed, the notebook's project directory is temporarily prepended to the module search path/

%Notebook-specific projects are named for the absolute path to the notebook file.
